---
title: "Gapminder Group Project"
author: "David Dvorak, Elton Chan, Fanli Si, Jingyang Liu"
date: "November 23, 2016"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
# Note that echo = FALSE hides all R code, change to TRUE to show R code in document
```

# Our Hypothesis:
Countries with higher per-capita GDP have higher life expectancies.

# Our Analysis:

```{r load_packages, message=FALSE, results='hide',include=FALSE}
require(ggplot2)
require(plyr)
require(dplyr)
require(gapminder)
#install.packages("gapminder")
```

As an overview let us consider how life expectancy varies as a function of per-capita GDP:

```{r}
df <- gapminder %>%
  mutate(gdpPercap_change = 100*((gdpPercap - lag(gdpPercap)))/gdpPercap) %>%
  mutate(lifeExp_change = 100*((lifeExp - lag(lifeExp)))/lifeExp) %>%
  filter(year > 1952)

qplot(gdpPercap,lifeExp,data=df, color = continent, xlab="Per-Capita GDP (int. dollars)", ylab="Life Expectancy (years)") + scale_x_log10()
qplot(gdpPercap,lifeExp,data=df, color = year, xlab="Per-Capita GDP (int. dollars)", ylab="Life Expectancy (years)") + scale_x_log10() + geom_smooth(method=lm, se = TRUE)
```

The results of the linear regression are as follows:

```{r}
fit_world <- lm(lifeExp ~ log(gdpPercap), data = df)
summary(fit_world)
```

And the residuals look like:

```{r}
qplot(predict(fit_world), residuals(fit_world),alpha=0.4,xlab="Predicted Life Expectancy",ylab="Residuals")
```

As a result we can say that there is a clear linear relationship between life expectancy and the logarithm of per-capita GDP.

To get a more clear picture about the relationship between life expectancy and per-capita GDP on different continents:

```{r}
ggplot(gapminder,
       aes(x = gdpPercap, y = lifeExp, group = continent, color = continent)) +
  geom_point(lwd = 1, show.legend = FALSE)  +  geom_smooth(method = "lm", formula = y ~ log(x)) + facet_wrap(~ continent) +
  scale_color_manual(values = continent_colors) +
  theme_bw() + theme(strip.text = element_text(size = rel(1.1))) +
  labs(x="Per-Capita GDP (int. dollars)", y="Life Expectancy (Years)")
```

It seems that GDP has a logarithmic relationship with life expectancy.

We can also look at how per-capita GDP and life expectancies have changed over time on each continent:

```{r}
ggplot(gapminder,
       aes(x = year, y = lifeExp, group = continent, color = continent)) +
  geom_point(lwd = 1, show.legend = FALSE)  +  geom_smooth(method="lm") + 
    facet_wrap(~ continent) + 
  scale_color_manual(values = continent_colors) +
  theme_bw() + theme(strip.text = element_text(size = rel(1.1))) +
  labs(x="Year", y="Life Expectancy (Years)")
  
ggplot(gapminder,
       aes(x = year, y = gdpPercap, group = continent, color = continent)) +
  geom_point(lwd = 1, show.legend = FALSE)  +  geom_smooth(method="lm") + facet_wrap(~ continent) +   scale_color_manual(values = continent_colors) +
  theme_bw() + theme(strip.text = element_text(size = rel(1.1))) +
  labs(x="Year", y="Per-Capita GDP (int. dollars)")

```

Interestingly, life expectancies across continents appear to increase a faster rate than per-capita GDP.

A follow-up question on the relationship between GDP and life expectancy is whether or not the rate of change of a country's life expectancy is impacted by the rate of change of its per-capita GDP. 

```{r}
ggplot(df, aes(x=gdpPercap_change,y=lifeExp_change, color = continent)) + geom_point() + geom_smooth(method=lm, se = FALSE) + ylim(-25,25) + xlim(-50,50) + labs(title="Comparing Growth Rates of Life Expectancy and Per-Capita GDP") +
  labs(x="Per-Capita GDP Growth Rate (%)", y="Life Expectancy Growth Rate (%)")
```

The scatterplot includes linear regression lines for each continent and the fitting results are as follows:

Africa:
```{r}
fit_africa <- lm(lifeExp_change ~ gdpPercap_change, data = (df %>% filter(continent == 'Africa')))
summary(fit_africa)$coef
```
The Americas:
```{r}
fit_americas <- lm(lifeExp_change ~ gdpPercap_change, data = (df %>% filter(continent == 'Americas')))
summary(fit_americas)$coef
```
Asia:
```{r}
fit_asia <- lm(lifeExp_change ~ gdpPercap_change, data = (df %>% filter(continent == 'Asia')))
summary(fit_asia)$coef
```
Europe:
```{r}
fit_europe <- lm(lifeExp_change ~ gdpPercap_change, data = (df %>% filter(continent == 'Europe')))
summary(fit_europe)$coef
```
Oceania:
```{r}
fit_oceania <- lm(lifeExp_change ~ gdpPercap_change, data = (df %>% filter(continent == 'Oceania')))
summary(fit_oceania)$coef
```

Note that the growth rate of per-capita GDP appears to have an impact on life expectancy growth rate for both Europe and Asia (p-value < 10<sup>-4</sup>), while having a minimal impact on the other continents.